Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features

### 2020-01-19 paper/paper.tex
fix minor bib issues; move Fig 3 to top of corresponding page

	%What we did briefly
	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware-dependent code optimization -- specifically by guiding the improvement of memory access patterns.
	We provide examples of manufacturer-recommended code optimizations to improve memory access patterns on the target architecture, and examine these using the architecture-independent workload characterization (AIWC)~\cite{beauaiwc} plugin for Oclgrind.
	Our work highlights the benefits and challenges arising from an architecture-independent analysis of memory-based program characteristics.
	We propose two new metrics to AIWC to better characterize these memory-based optimizations.
	We measure the presented codes using the new metrics and demonstrate the metrics' effectiveness in capturing the essence of the optimizations performed.

@@

	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
	Global memory is typically located off-chip and accesses induce large delays.
	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory \cite{cudamanual}. Appendix \ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).

@@

@@

	\begin{figure*}[h!]
		\centering
		\subfloat{
			\centering
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT,alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
			legend style={at={(0.6,0.45)},anchor=west}
			]

@@

	\begin{table*}[h!]
	\centering
	\begin{tabular}[t]{lrrrrr}
		\toprule
		& simple & coalescedA & coalescedAB & coalescedABT & alignedABT \\ \midrule 
		Total memory footprint        & 196608 & 196608     & 196608      & 196608       & 196608     \\ %\hline
		90\% Memory Footprint         & 118196 & 56176      & 489         & 489          & 489        \\ %\hline
		Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78         & 9.78       \\ %\hline
		LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78         & 8.78       \\ %\hline
		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78         & 1.78       \\ %\hline
		Relative Local Memory Usage   & 0      & 49.91    & 94.28       & 94.28        & 94.28      \\ \bottomrule
	\end{tabular}
	\caption{Selection of AIWC\cite{beauaiwc} metrics for 256 $\times$ 256 matrix multiplication}
	\label{matmul aiwc table}
	\end{table*}

@@

	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simpleMultiply} to \texttt{coalescedAB}, we find that the ratio of OpenCL local memory usage increases as expected.

>>>>

	%What we did briefly
	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware-dependent code optimization -- specifically by guiding the improvement of memory access patterns.
	We provide examples of manufacturer-recommended code optimizations to improve memory access patterns on the target architecture, and examine these using the architecture-independent workload characterization (AIWC)~\cite{beauaiwc} plugin for Oclgrind~\cite{price:15}.
	Our work highlights the benefits and challenges arising from an architecture-independent analysis of memory-based program characteristics.
	We propose two new metrics to AIWC to better characterize these memory-based optimizations.
	We measure the presented codes using the new metrics and demonstrate the metrics' effectiveness in capturing the essence of the optimizations performed.

@@

	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
	Global memory is typically located off-chip and accesses induce large delays.
	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).

@@

	\begin{figure*}[ht!]
		\centering
		\subfloat{
			\centering
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT,alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
			legend style={at={(0.6,0.45)},anchor=west}
			]

@@

@@

	\begin{table*}[h!]
	\centering
	\begin{tabular}[t]{lrrrrr}
		\toprule
		& simple & coalescedA & coalescedAB & coalescedABT & alignedABT \\ \midrule 
		Total memory footprint        & 196608 & 196608     & 196608      & 196608       & 196608     \\ %\hline
		90\% Memory Footprint         & 118196 & 56176      & 489         & 489          & 489        \\ %\hline
		Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78         & 9.78       \\ %\hline
		LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78         & 8.78       \\ %\hline
		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78         & 1.78       \\ %\hline
		Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94         & 0.94      \\ \bottomrule
	\end{tabular}
	\caption{Selection of AIWC\cite{beauaiwc} metrics for 256 $\times$ 256 matrix multiplication}
	\label{matmul aiwc table}
	\end{table*}

@@

	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simpleMultiply} to \texttt{coalescedAB}, we find that the proportion of OpenCL local memory usage increases as expected.
